# Bacterial and fungal community analysis

lorem ipsum

First we need to load some libraries.  

```{r}
.libPaths(c("/projappl/project_2007145/project_rpackages_r421", .libPaths()))
libpath <- .libPaths()[1]

library(phyloseq)
library(microViz)
library(tidyverse)
library(vegan)
```

Read in the data. 

```{r}
bac_raw <- readRDS("outputs/physeq_bac.rds")
fun_raw <- readRDS("outputs/physeq_fun.rds")

bac_raw <- bac_raw %>% subset_taxa(Kingdom!="NA")
fun_raw <- fun_raw %>% subset_taxa(Kingdom!="NA")
```

First we'll plot some metadata parameters.  
As they are the same for both fungi and bacteria, we can take the metadata table from the bacteria phyloseq object.

```{r}
metadata <- sample_data(bac_raw) %>%
                data.frame() %>%
                rownames_to_column("sample_name") %>%
                tibble() 
metadata
```

We have the negative controls in there, so we need to get rif off them. And also get rid of the factor 

```{r}
sample_metadata <- metadata %>% 
                filter(Site!="NA") %>% droplevels()
sample_metadata
```

The quick and dirty way of looking at your data (we can leave out the `Group` variable).

```{r}
plot(sample_metadata[2:8])
```


Then we could look at the differences in pH between park, forest and agricultual soils. 

```{r}
sample_metadata %>% 
        ggplot(aes(x=Site, y=pH, fill=Site)) + geom_boxplot()
```

What else could be interesting to plot? Change the `x` and `y`.  

```{r}
sample_metadata %>% 
        ggplot(aes(x=XX, y=YY, fill=Site)) + geom_boxplot()
```

Then it's time to look at the microbial community data.  
We'll use the bacterial data as an example and then you can explore the fundal data on your own. 

Let's start with some preliminary data exploration

First the total count per sample (sample size, library size...)

```{r}
bac_raw %>% sample_sums()
```

We can  also plot that 

```{r}
bac_raw %>% 
    sample_sums() %>% 
    barplot(las=3)
```

Then using the `ord_explore` function.

```{r}
bac_raw %>% 
    tax_fix() %>%
    ord_explore
```


We probably need to remove some samples from our data set. At least the negative controls. 
```{r}
bac_raw <- bac_raw %>% 
                subset_samples(Site!="NA")

bac_raw <- bac_raw %>% 
                prune_samples(sample_names(.)!="A026-D-fore-bac", .)
```

Also remember to remove the sample from the metadata.

```{r}
sample_metadata <- sample_metadata %>% subset(sample_name != "A026-D-fore-bac")
```

We can make two different versions of our data.  
One with relative abundances (proportions) and one with centered log-ratio transformation.

```{r}
bac_ra <- bac_raw %>% transform_sample_counts(function(x) x/sum(x))
bac_clr <- bac_raw %>% tax_transform(trans="clr")
```

Barplot of taxa relative abundances
```{r}
bac_ra %>% 
    tax_fix() %>%
    comp_barplot("Genus", n_taxa = 19, merge_other = TRUE, label = "Group") + 
        facet_grid(.~Site, scales="free")
```

Now do on a different taxonomic level.

```{r}
bac_ra %>% 
    tax_fix() %>%
    comp_barplot()
```

Alpha diversity measures.  
For that we need the species table from our phyloseq object.  
We will use the diversity functions in vegan, which expects taxa to be columns by default. So we can either transpose the matrix or use the `MARGIN` option in the functions.

```{r}
asv_table_raw <- otu_table(bac_raw) %>% t()
asv_table_ra <- otu_table(bac_ra) %>% t()
asv_table_clr <- otu_table(bac_clr) %>% t()
```

First we can count the number of observed "species" (ASVs). 

```{r}
asv_table_raw %>% 
    specnumber(MARGIN=1)
```

And also plot that.

```{r}
asv_table_raw %>% 
    specnumber(MARGIN=1) %>% 
    data.frame(index=.) %>% 
    rownames_to_column("sample_name") %>%
    left_join(., metadata) %>% 
    ggplot(aes(x=Site, y=index, fill=Site)) + 
        geom_boxplot() +
        labs(y="Observed ASVs", x="")
```

Is the number of observed species dependent on the library size? Maybe we should check. 

```{r}
obs_asv <- asv_table_raw %>% 
    specnumber(MARGIN=1) %>% 
    data.frame(obs=.) %>% 
    rownames_to_column("sample_name")

total_count <- asv_table_raw %>% 
    sample_sums() %>% 
    data.frame(sum=.) %>% 
    rownames_to_column("sample_name")

left_join(obs_asv, total_count) %>% ggplot(aes(x=sum, y=obs)) + geom_point()
```

To get comparable values, we nee dto rarefy the data. 
```{r}
bac_rf <- rarefy_even_depth(bac_raw)
asv_table_rf <- otu_table(bac_rf)

asv_table_rf %>% 
    specnumber(MARGIN=1) %>% 
    data.frame(index=.) %>% 
    rownames_to_column("sample_name") %>%
    left_join(., metadata) %>% 
    ggplot(aes(x=Site, y=index, fill=Site)) + 
        geom_boxplot() +
        labs(y="Observed ASVs (rarefied data)", x="")
```

Then we can calculate the shannon index from the raw data.

```{r}
asv_table_raw %>% 
    diversity(index="shannon", MARGIN=1)
```

And again also plot that.

```{r}
asv_table_raw %>% 
    diversity(index="shannon", MARGIN=1) %>% 
    data.frame(index=.) %>% 
    rownames_to_column("sample_name") %>%
    left_join(., metadata) %>% 
    ggplot(aes(x=Site, y=index, fill=Site)) + 
        geom_boxplot() +
        labs(y="Shannon index", x="")
```

Now do the same using the proportions and with simpson's index (`index="simpson"`).

```{r}
SOME_CODE_HERE
```

Beta-diversity and ordinations.   

First we need distance matrices calculated from the normalized/transformed data.
We'll use the Bray-Curtis dissimilarity index for the relative abundances and euclidean distances for the clr transformed data.

```{r}
dist_ra <- asv_table_ra %>% vegdist(method="bray") 
dist_clr <- asv_table_clr %>% vegdist(method="euclidean") 
```

Then we can plot a PCoA plot from the Bray-Curtis dissimilarities. 
```{r}
dist_ra %>% 
    cmdscale() %>%  
    data.frame() %>% 
    rownames_to_column("sample_name") %>%
    left_join(., sample_metadata) %>% 
    ggplot(aes(x=X1, y=X2, color=Site)) + geom_point() 
```

And the same for the clr/euclidean data.

```{r}
dist_clr %>%
    cmdscale() %>%  
    data.frame() %>% 
    rownames_to_column("sample_name") %>%
    left_join(., sample_metadata) %>% 
    ggplot(aes(x=X1, y=X2, color=Site)) + geom_point() 
```

Hypothesis testing. 

```{r}
SOME_CODE_HERE

```
