# Data Exploration


## Libraries

```{r}
library(reshape2)
library(dplyr)
library(tidyr)
library(MASS)
library(vegan)
library(ggplot2)
library(ggthemes)
library(colorspace)
library(colorRamps)
library(RColorBrewer)
library(gridExtra)
library(fitdistrplus)
library(logspline)
library(car)
library(lme4)
# library(multcomp)
library(scales)
library(viridis)
library(ggrepel)
library(pheatmap)
# library(psych)
```


## Setup

```{r}
setwd("PATH/TO/COURSE/FOLDER")
```

Read in matadata
```{r}
MMB117metadata <- read.table("MMB117_Sample_sheet_2025.txt", header = TRUE, stringsAsFactors = TRUE, sep = "\t")

View(MMB117metadata)
MMB117metadata <- MMB117metadata[1:25, 1:20]
```

## Read in data

```{r}
ASV_table <- readRDS("seqtab.rds")

str(ASV_table)
View(ASV_table)

tax_table <- readRDS("taxa.rds")
str(tax_table)

View(tax_table)

identical(rownames(tax_table), colnames(ASV_table))
# [1] TRUE

head(rownames(tax_table))

head(colnames(ASV_table))

colnames(ASV_table) <- paste0("ASV", seq(ncol(ASV_table)))

rownames(tax_table) <- paste0("ASV", seq(nrow(tax_table)))

# I will store the row sums for later use if I need:
ASV_counts_per_sampleDF <- as.data.frame(rowSums(ASV_table))

# I will check if there are ASVs that are detected only less than 3 times

dim(ASV_table)
# [1]    26 20262

dim(ASV_table[, which(colSums(ASV_table) < 3)])
# [1]  26 606

# One could remove these

# ASV_table_over3<-ASV_table[, which(colSums(ASV_table)>3)]

str(ASV_table)
class(ASV_table)
```


## TSS normalization to get realtive abundances for data exploration

```{r}
ASV_tableRA.mat<- apply(ASV_table, 1, function(i) i/sum(i))
ASV_tableRA.mat<-t(ASV_tableRA.mat)

View(ASV_tableRA.mat)

# I will remove the negative control from BI

ASV_tableRA.mat<-ASV_tableRA.mat[2:nrow(ASV_tableRA.mat),]
View(MMB117metadata)

# # the Sample names are different...
# # I will change them.

# MMB117metadata$Sample
# #[1] GA_01        GA_02        GA_03        GA_04        GA_05        GA_06        PA_01        PA_02        PA_03        PA_04        PA_05        PA_06       
# #[13] FI_01        FI_02        FI_03        FI_04        FI_05        FI_06        FO_01        FO_02        FO_03        FO_04        FO_05        FO_06       
# #[25] Neg. control

# rownames(ASV_tableRA.mat)
# #[1] "CTRL-TATCCAGT" "FI-01"         "FI-02"         "FI-03"         "FI-04"         "FI-05"         "FI-06"         "FO-01"         "FO-02"         "FO-03"        
# #[11] "FO-04"         "FO-05"         "FO-06"         "GA-01"         "GA-02"         "GA-03"         "GA-04"         "GA-05"         "GA-06"         "PA-01"        
# #[21] "PA-02"         "PA-03"         "PA-04"         "PA-05"         "PA-06"        

# MMB117Samples<- c("Neg. control", "FI_01", "FI_02", "FI_03",  "FI_04",  "FI_05",  "FI_06",  
#                   "FO_01" , "FO_02", "FO_03", "FO_04",  "FO_05" , "FO_06",
#                   "GA_01",  "GA_02" ,  "GA_03",  "GA_04",  "GA_05", "GA_06",
#                   "PA_01",  "PA_02" , "PA_03" ,"PA_04"  ,"PA_05",   "PA_06")
# length(MMB117Samples)
# #[1] 25

# rownames(ASV_tableRA.mat)<-MMB117Samples

# a basic NMDS plot usually tells if there is something funny with the data

plot(metaMDS(ASV_tableRA.mat), type="text", display="sites")

#Warning message:
#In metaMDS(ASV_tableRA.mat) :
 # stress is (nearly) zero: you may have insufficient data

# because of the negative control, as the plot shows.

plot(metaMDS(ASV_tableRA.mat[2:25,]), type="text", display="sites")

# looks pretty good!

# For the future I want to put the samples in the same order in the metadata and ASV data.

ASV_tableRA.mat_o<-ASV_tableRA.mat[match(MMB117metadata$Sample,row.names(ASV_tableRA.mat)),]

# Now I can also get Shannons diversity index and put that into metadata

MMB117metadata$ASV_divSha<-diversity(ASV_tableRA.mat_o)




###################################################################################

#                                 Data exploration                                #

###################################################################################

# According to
# Zuur, A.F., Ieno, E.N. and Elphick, C.S. (2010), 
# A protocol for data exploration to avoid common statistical problems. 
# Methods in Ecology and Evolution, 1: 3-14. https://doi.org/10.1111/j.2041-210X.2009.00001.x

#############################################################################################


# I want to start from ***INDEPENDENCE***, since it is the most important assumption:

# We had only one sampling point.
# Thus we don't really have "traditional" independence problem which would arise from repeated measurements.
# If we would have taken samples before and after some treatment from the same 
# sites, we would have to take this into account in our statistical analyses
# (use mixed models, etc.)

############################################################################################

# --- Outliers in X & Y--- #

# X = metadata and things that are usually on x-axis

# Y = the numerical data we use, in this case we have  microbiome data that are either counts (non-normalized), 
# relative abundance (TSS-normalized), or clr-transformed normally distributed numbers that can be used 
# in statistical analyses

# metadata first.
str(MMB117metadata)

#'data.frame':	25 obs. of  20 variables:
# $ Sample                  : Factor w/ 26 levels "","FI_01","FI_02",..: 14 15 16 17 18 19 21 22 23 24 ...
#$ Site                    : Factor w/ 6 levels "","-","Field",..: 5 5 5 5 5 5 6 6 6 6 ...
#$ SnowDepth               : Factor w/ 6 levels "","0 cm","10 cm",..: 3 3 1 1 2 1 4 4 1 3 ...
#$ WetWeight               : num  10 10 11.2 17.4 10 ...
#$ DryWeight               : num  6.51 6.47 6.84 13.51 5.57 ...
#$ X80C_DryWeight          : num  NA NA 6.79 13.43 5.44 ...
#$ DNAWeight               : num  306 237 501 285 340 ...
#$ Moisture                : num  58 58 65.1 32 84.7 ...
#$ pH_W                    : num  6.61 5.76 6.18 6.37 6.19 6.37 5.8 6.37 6.6 6.4 ...
#$ pH_Ca                   : num  5.63 5.24 5.72 5.77 5.58 5.62 5.34 5.57 5.81 5.74 ...
#$ SOM                     : num  13.5 12 7.6 3.8 17.6 ...
#$ ResponsiblePerson       : Factor w/ 10 levels "","Arttu","Erin",..: 4 4 6 6 5 5 9 9 9 3 ...
#$ Notes                   : Factor w/ 21 levels "","clay soil with too much water",..: 7 12 14 20 6 19 3 1 9 13 ...
#$ X10_5_TSA_Plate         : Factor w/ 17 levels "","0","1","103",..: 10 4 12 5 14 17 15 11 6 13 ...
#$ X10_6_TSA_Plate         : int  1 12 1 1 3 5 6 2 4 10 ...
#$ X10_7_TSA_Plate         : int  0 3 3 0 0 1 0 0 0 1 ...
#$ X10_4_Malt_Plate        : int  18 30 11 7 17 12 1 2 0 8 ...
#$ X10_5_Malt_Plate        : int  8 6 1 1 3 1 0 0 0 2 ...
#$ X10_6_Malt_Plate        : int  0 1 0 0 1 0 0 0 0 1 ...
#$ Other_plate_observations: Factor w/ 6 levels "","10-5 malt plate had fungal growth",..: 1 1 1 1 2 5 1 1 1 1 ...
#$ ASV_divSha              : num  7.19 7.33 7.2 7.07 7.27 ...


# So lets plot all variables that make sense to look at 

ggplot(MMB117metadata, aes(Sample, SnowDepth, label=Site)) +
 geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Snow depth")

# not recorded for all, maybe can be left put from the analysis

ggplot(MMB117metadata, aes(Sample,  WetWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Wet weight (g)")

#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control


ggplot(MMB117metadata, aes(Sample,  DryWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Dry weight (air dried) (g)")

#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control


ggplot(MMB117metadata, aes(Sample,  X80C_DryWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Owen dry weight, pore water dried out (g)")

#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control


ggplot(MMB117metadata, aes(Sample,  DNAWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
ggtitle("What was weighed for DNA extraction (mg)")

#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control


ggplot(MMB117metadata, aes(Sample,  Moisture, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Soil water % (?)")

#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control

#Gravimetric water content would be better than "moisture", you could change the column. 
#There are some that are very wet. 
#Interesting to see if their microbiome is completely different because of the snow for instance


ggplot(MMB117metadata, aes(Sample,  pH_W, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Water-extraxted pH")


#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control


ggplot(MMB117metadata, aes(Sample,  pH_Ca, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("CaCl2-extraxted pH")

# nice grouping!!

# With both pH's actually

#Warning messages:
#1: Removed 1 row containing missing values or values outside the scale range (`geom_point()`). 
#2: Removed 1 row containing missing values or values outside the scale range (`geom_text_repel()`). 

# negative control


ggplot(MMB117metadata, aes(Sample,  SOM, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Soil organic matter (LOI)")

# few samples that are under or above but not too extreme I would say
# some missing data as well


ggplot(MMB117metadata, aes(Sample,  ASV_divSha, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Shannon's diversity index, Y data")

# negative control is an outlier, as it should

# The NMDS plot also shows if there are otliers in the samples (in terms of Y data)
plot(metaMDS(ASV_tableRA.mat[2:25,]), type="text", display="sites")
# and not really

plot(metaMDS(ASV_tableRA.mat[2:25,]))

# this shows both the species and sites and from this we cans see that different 
# species associate to different sites

              ##----------------------------- ##

# Good! looks like there are no weird extreme cases in the data.

###################################################################

# --- Homogenity of variances --- #

# With statistical analyses we usually compare means, so if the black lines are in different levels,
# that is fine. 
# but the variability of the observations should be similar. 
# i.e. the size of the boxes should be about the same

# So lets check this with metadata and Shannons diversity index

ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  Moisture)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")


# whoops! Good thing we are not studying change of soil water % in response to land use!
# but I wouldn't build a linear model using soil water % as an explaining variable.
# non-parametric statistical test would be probably ok.


ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  pH_W)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")

# same here but less extreme

ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  pH_Ca)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")

# this could work with liner model as an X-variable

ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  SOM)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")

#Warning message:
#Removed 3 rows containing non-finite outside the scale range (`stat_boxplot()`). 

# Some samples dont have SOM -values

# Gas station is "in another planet", so SOM as an X-variable of a linear model would probably be a bad idea.

ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  ASV_divSha)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")


# Also diversity shows heterogeneity in the variances, but this probably would work in linear models as an Y-variable

                          ##----------------------------- ##

# We can take these into account by selecting statistical models that that do not require homogeneity.
# If we would really need to use linear models, we could also transform these variables
# or delete observations that are outliers and go with for instance 3 observations per site.
# This is the reason why even 6 observations (6 samples) per site can bee too little!


###############################################################################################

# --- Normality --- #

#Often people say that the data should be normally distributed, 
#if you want to apply linear models.  
#This is not exactly true: The normality should be met at each X value. 
#This is of difficult to check if the data does not have multiple 
#observations for each X and we don't. 
#In other words, the Y-data contains the effects of all the explanatory variables (all metadata).
#Therefore it is misleading to assess normality according the Y data.
#Better choice is to apply a model and plot the pooled residuals. 
#Since the linear models should be LINEAR, the residuals should have 
#the same distance to fitted values (predicted means). If they do not, 
#then the model is not linear and the plot of residuals against fitted 
#values will show a curve or some other pattern. This means that 
#the RESIDUALS should be normally distributed, if linear models are used.

#We know this about normal distribution:
# "It is a continuous probability distribution in which the random variable can take any value." 
#With TSS-normalization, our values are "relative abundances", so something between 0 and 1.
#So relative abundances don't meet the normality assumption and we cannot use a model that 
#assumes normality with TSS-normalized data.

# Relative abundances are proportions, so the probability distribution to use with them would be 
# Beta distribution, but according to my experience, Gamma distribution works also 
# Gamma distribution is used to model positive continuous variables that represent time intervals between events, 
# the Beta distribution is used to model continuous variables that represent proportions or probabilities
# Gamma models work very well, I have had trouble with Beta

#######################################################################################################
# But what should be normally distributed even though it would be calculated from relative abundances:
# Shannon's diversity index
#####################################################################################################
#####################################################################################################
# Some metadata variables (that are not %) would be normally distributed, for instance pH.
#####################################################################################################

# tehäänkö tähän liittyen jotain? äänestän et ei. Tää voi mennä teorialla, menee liian vaikeeksi muuten ja ohi asian

# ---  "Fixed X" --- #

#This assumption means that you know the values of X for each sample in advance. 
#You can have e.g. different fertilization rates, treatments, etc. 
#This assumption is violated, if you for instance study some parameter of animals,
#which happen to be found somewhere and your X is for example the age of animals. In other words, 
#this assumption is violated if the X is randomly selected.

# we don't have this problem

###########################################################################################################

# ---  Zero trouble in  Y-data  --- #

# if most observations are zeros

# we probable have many zeros, at least in the ASV and genera levels

range(ASV_tableRA.mat_o)
# [1] 0.0000000 0.1498224

plot(c(ASV_tableRA.mat_o))

# and it seems we do. So if we would like to apply linear models using taxa as the response variables,
# we might need to consider zero-inflated models

###########################################################################################################

# --- Collinearity in X and # --- Relationships Y & X --- #

#If X-observations correlate, for instance weight and length, or water depth and distance to the shoreline. 
#If collinearity is ignored, one is likely to end up with a confusing statistical analysis in which nothing 
#is significant, but where dropping one covariate can make the others significant, 
#or even change the sign of estimated parameters.

# collinearity can be detected with pairwise scatter plots comparing covariates,


plot(MMB117metadata[,c(8:11)])

# pH water and pH CaCl2 seem to correlate, as they should. But because of the collinearity issue both of them cannot be used in a model

# relationships: Y-values should be plotted against X-values to see which X-values should be included in the model. 

plot(MMB117metadata[,c(8:11,21)])

# This was a lazy persons solution. We can see that these is a relationship with the diversity and the X-variables. 
# And this also makes bilogically sense.

# So if we would do linear models and use diversity as the response variable, we might need to consider
# water %, pH (one of them) and SOM %

###########################################################################################################

# --- Interactions? --- #
#

# This means that if we would do linear models, some of the X-variables might have an interaction and this can be taken into account

# Let's try to visualize if we have something like this.

# I will categorize water-content

summary(MMB117metadata$Moisture)

#Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
# 29.24   38.50   42.16   47.19   54.97   84.68       1 

Water_FirstQ<-38.50
Water_Median<-42.16 
Water_ThirdQ<- 54.97

MMB117metadata$MoistureCAT<- cut(MMB117metadata$Moisture,c(0, 45,  84.68  ))

MMB117metadata$MoistureCAT  

levels(MMB117metadata$MoistureCAT )  
#[1] "(0,45]"    "(45,84.7]"

levels(MMB117metadata$MoistureCAT )  <- c("0 - 45 %","45 - 84.7 %")

MMB117metadata$MoistureCAT


ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(x=pH_Ca, y=ASV_divSha, colour=Site)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  scale_color_manual(values=c("springgreen4", "chocolate4","purple4","royalblue"))+
  facet_grid(Site ~ MoistureCAT) +
  theme_bw(base_size = 14)+
  theme(strip.background = element_rect(colour = "black", fill = "white"))+
  theme(panel.border = element_rect(colour = "black"))+
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())+
  #guides(color=guide_legend(title="Sample type"))+
  theme(legend.position="none")

# Looks like in our case we don't have enough observations for this interaction