---
title: MMB-117
output: 
        github_document:
        toc: true
        toc_float: true
        number_sections: true
        fig_caption: true
        keep_md: true
        highlight: tango
        theme: united
        dev: svg
        self_contained: no
---

# Data Exploration

## Libraries

```{r "libraries"}
library(tidyverse)
# library(reshape2)
# library(dplyr)
# library(tidyr)
# library(MASS)
# library(vegan)
# library(ggplot2)
# library(ggthemes)
# library(colorspace)
# library(colorRamps)
# library(RColorBrewer)
# library(gridExtra)
# library(fitdistrplus)
# library(logspline)
# library(car)
# library(lme4)
# # library(multcomp)
# library(scales)
# library(viridis)
# library(ggrepel)
# library(pheatmap)
# # library(psych)
```


## Setup

```{r}
setwd("PATH/TO/COURSE/FOLDER")
```

Read in the phyloseq object and extract the metadata from the object.  

```{r "read_metadata"}
physeq <- readRDS("physeq.rds")
MMB117metadata <- sample_data(physeq)

View(MMB117metadata)
```

Read in data

```{r}
ASV_table <- physeq %>%
  otu_table() %>%
  as.data.frame() %>%
  t()

str(ASV_table)
View(ASV_table)

tax_table <- physeq %>%
  tax_table() %>%
  as.data.frame()
str(tax_table)

View(tax_table)

identical(rownames(tax_table), colnames(ASV_table))
# [1] TRUE

head(rownames(tax_table))

head(colnames(ASV_table))

colnames(ASV_table) <- paste0("ASV", seq(ncol(ASV_table)))

rownames(tax_table) <- paste0("ASV", seq(nrow(tax_table)))

# I will store the row sums for later use if I need:
ASV_counts_per_sampleDF <- as.data.frame(rowSums(ASV_table))

# I will check if there are ASVs that are detected only less than 3 times

dim(ASV_table)
# [1]    26 20262

dim(ASV_table[, which(colSums(ASV_table) < 3)])
# [1]  26 606

# One could remove these

# ASV_table_over3<-ASV_table[, which(colSums(ASV_table)>3)]

str(ASV_table)
class(ASV_table)
```


## TSS normalization to get realtive abundances for data exploration

```{r}
ASV_tableRA.mat<- apply(ASV_table, 1, function(i) i/sum(i))
ASV_tableRA.mat<-t(ASV_tableRA.mat)

View(ASV_tableRA.mat)

# I will remove the negative control from BI

ASV_tableRA.mat<-ASV_tableRA.mat[2:nrow(ASV_tableRA.mat),]
View(MMB117metadata)

# # the Sample names are different...
# # I will change them.

# MMB117metadata$Sample
# #[1] GA_01        GA_02        GA_03        GA_04        GA_05        GA_06        PA_01        PA_02        PA_03        PA_04        PA_05        PA_06       
# #[13] FI_01        FI_02        FI_03        FI_04        FI_05        FI_06        FO_01        FO_02        FO_03        FO_04        FO_05        FO_06       
# #[25] Neg. control

# rownames(ASV_tableRA.mat)
# #[1] "CTRL-TATCCAGT" "FI-01"         "FI-02"         "FI-03"         "FI-04"         "FI-05"         "FI-06"         "FO-01"         "FO-02"         "FO-03"        
# #[11] "FO-04"         "FO-05"         "FO-06"         "GA-01"         "GA-02"         "GA-03"         "GA-04"         "GA-05"         "GA-06"         "PA-01"        
# #[21] "PA-02"         "PA-03"         "PA-04"         "PA-05"         "PA-06"        

# MMB117Samples<- c("Neg. control", "FI_01", "FI_02", "FI_03",  "FI_04",  "FI_05",  "FI_06",  
#                   "FO_01" , "FO_02", "FO_03", "FO_04",  "FO_05" , "FO_06",
#                   "GA_01",  "GA_02" ,  "GA_03",  "GA_04",  "GA_05", "GA_06",
#                   "PA_01",  "PA_02" , "PA_03" ,"PA_04"  ,"PA_05",   "PA_06")
# length(MMB117Samples)
# #[1] 25

# rownames(ASV_tableRA.mat)<-MMB117Samples

# a basic NMDS plot usually tells if there is something funny with the data

plot(metaMDS(ASV_tableRA.mat), type="text", display="sites")

#Warning message:
#In metaMDS(ASV_tableRA.mat) :
 # stress is (nearly) zero: you may have insufficient data

# because of the negative control, as the plot shows.

plot(metaMDS(ASV_tableRA.mat[2:25,]), type="text", display="sites")

# looks pretty good!

# For the future I want to put the samples in the same order in the metadata and ASV data.

ASV_tableRA.mat_o<-ASV_tableRA.mat[match(MMB117metadata$Sample,row.names(ASV_tableRA.mat)),]

# Now I can also get Shannons diversity index and put that into metadata

MMB117metadata$ASV_divSha<-diversity(ASV_tableRA.mat_o)
```


## Data exploration

According to:
Zuur, A.F., Ieno, E.N. and Elphick, C.S. (2010), 
A protocol for data exploration to avoid common statistical problems. 
[Methods in Ecology and Evolution, 1: 3-14](https://doi.org/10.1111/j.2041-210X.2009.00001.x)

I want to start from ***INDEPENDENCE***, since it is the most important assumption:

We had only one sampling point. Thus we don't really have "traditional" independence problem which would arise from repeated measurements.  
If we would have taken samples before and after some treatment from the same sites, we would have to take this into account in our statistical analyses
(use mixed models, etc.)

### Outliers in X & Y-

X = metadata and things that are usually on x-axis

Y = the numerical data we use, in this case we have  microbiome data that are either counts (non-normalized),  
relative abundance (TSS-normalized), or clr-transformed normally distributed numbers that can be used in statistical analyses

Metadata first.

```{r}
str(MMB117metadata)
```


So lets plot all variables that make sense to look at 

```{r}
ggplot(MMB117metadata, aes(Sample, SnowDepth, label=Site)) +
 geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Snow depth")
```

Not recorded for all, maybe can be left out from the analysis

```{r}
ggplot(MMB117metadata, aes(Sample,  WetWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Wet weight (g)")

ggplot(MMB117metadata, aes(Sample,  DryWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Dry weight (air dried) (g)")

ggplot(MMB117metadata, aes(Sample,  X80C_DryWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Owen dry weight, pore water dried out (g)")

```

```{r}
ggplot(MMB117metadata, aes(Sample,  DNAWeight, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
ggtitle("What was weighed for DNA extraction (mg)")


ggplot(MMB117metadata, aes(Sample,  Moisture, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Soil water % (?)")
```


Gravimetric water content would be better than "moisture", you could change the column. 
There are some that are very wet. 
Interesting to see if their microbiome is completely different because of the snow for instance

```{r}
ggplot(MMB117metadata, aes(Sample,  pH_W, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Water-extraxted pH")

ggplot(MMB117metadata, aes(Sample,  pH_Ca, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("CaCl2-extraxted pH")
```

Nice grouping!!

With both pH's actually

```{r}
ggplot(MMB117metadata, aes(Sample,  SOM, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Soil organic matter (LOI)")
```

Few samples that are under or above but not too extreme I would say
Some missing data as well

```{r}
ggplot(MMB117metadata, aes(Sample,  ASV_divSha, label=Site)) +
  geom_point(color="black", size=3)+
  theme(axis.text.x = element_text(colour = "black", size = 8, angle = 270, hjust=0, vjust=0.5))+
  geom_text_repel(color="black",size=4,max.overlaps = Inf )+
  ggtitle("Shannon's diversity index, Y data")

# negative control is an outlier, as it should

# The NMDS plot also shows if there are otliers in the samples (in terms of Y data)
plot(metaMDS(ASV_tableRA.mat[2:25,]), type="text", display="sites")
# and not really

plot(metaMDS(ASV_tableRA.mat[2:25,]))
```

This shows both the species and sites and from this we cans see that different species associate to different sites

__Good! looks like there are no weird extreme cases in the data.__

## Homogenity of variances

With statistical analyses we usually compare means, so if the black lines are in different levels, that is fine.  
But the variability of the observations should be similar. i.e. the size of the boxes should be about the same

So lets check this with metadata and Shannons diversity index

```{r}
ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  Moisture)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")
```

whoops! Good thing we are not studying change of soil water % in response to land use!
But I wouldn't build a linear model using soil water % as an explaining variable. non-parametric statistical test would be probably ok.

```{r}
ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  pH_W)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")

# same here but less extreme

ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  pH_Ca)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")

# this could work with liner model as an X-variable

ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  SOM)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")
```

Some samples dont have SOM -values

Gas station is "in another planet", so SOM as an X-variable of a linear model would probably be a bad idea.

```{r}
ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(Site,  ASV_divSha)) + 
  geom_boxplot() +
  theme_bw(base_size = 14)+
  theme(panel.border = element_rect(colour = "black"))+
  theme(legend.position="none")
```

Also diversity shows heterogeneity in the variances, but this probably would work in linear models as an Y-variable
We can take these into account by selecting statistical models that that do not require homogeneity.
If we would really need to use linear models, we could also transform these variables
or delete observations that are outliers and go with for instance 3 observations per site.
This is the reason why even 6 observations (6 samples) per site can bee too little!


### Normality

Often people say that the data should be normally distributed, 
if you want to apply linear models.  
This is not exactly true: The normality should be met at each X value. 
This is of difficult to check if the data does not have multiple 
observations for each X and we don't. 
In other words, the Y-data contains the effects of all the explanatory variables (all metadata).
Therefore it is misleading to assess normality according the Y data.
Better choice is to apply a model and plot the pooled residuals. 
Since the linear models should be LINEAR, the residuals should have 
the same distance to fitted values (predicted means). If they do not, 
then the model is not linear and the plot of residuals against fitted 
values will show a curve or some other pattern. This means that 
the RESIDUALS should be normally distributed, if linear models are used.

We know this about normal distribution:
 "It is a continuous probability distribution in which the random variable can take any value." 
With TSS-normalization, our values are "relative abundances", so something between 0 and 1.
So relative abundances don't meet the normality assumption and we cannot use a model that 
assumes normality with TSS-normalized data.

Relative abundances are proportions, so the probability distribution to use with them would be 
Beta distribution, but according to my experience, Gamma distribution works also 
Gamma distribution is used to model positive continuous variables that represent time intervals between events, 
the Beta distribution is used to model continuous variables that represent proportions or probabilities
Gamma models work very well, I have had trouble with Beta


But what should be normally distributed even though it would be calculated from relative abundances:
Shannon's diversity index
Some metadata variables (that are not %) would be normally distributed, for instance pH.

### Fixed X

This assumption means that you know the values of X for each sample in advance. 
You can have e.g. different fertilization rates, treatments, etc. 
This assumption is violated, if you for instance study some parameter of animals,
which happen to be found somewhere and your X is for example the age of animals. In other words, 
this assumption is violated if the X is randomly selected.

we don't have this problem.

### Zero trouble in  Y-data

if most observations are zeros
we probable have many zeros, at least in the ASV and genera levels

```{r}
range(ASV_tableRA.mat_o)
# [1] 0.0000000 0.1498224

plot(c(ASV_tableRA.mat_o))

# and it seems we do. So if we would like to apply linear models using taxa as the response variables,
# we might need to consider zero-inflated models
```


### Collinearity in X and relationships Y & X

If X-observations correlate, for instance weight and length, or water depth and distance to the shoreline. 
If collinearity is ignored, one is likely to end up with a confusing statistical analysis in which nothing 
is significant, but where dropping one covariate can make the others significant, 
or even change the sign of estimated parameters.

collinearity can be detected with pairwise scatter plots comparing covariates,

```{r}
plot(MMB117metadata[,c(8:11)])

# pH water and pH CaCl2 seem to correlate, as they should. But because of the collinearity issue both of them cannot be used in a model

# relationships: Y-values should be plotted against X-values to see which X-values should be included in the model. 

plot(MMB117metadata[,c(8:11,21)])

# This was a lazy persons solution. We can see that these is a relationship with the diversity and the X-variables. 
# And this also makes bilogically sense.

# So if we would do linear models and use diversity as the response variable, we might need to consider
# water %, pH (one of them) and SOM %
```

### Interactions?


This means that if we would do linear models, some of the X-variables might have an interaction and this can be taken into account

Let's try to visualize if we have something like this.

I will categorize water-content

```{r}
summary(MMB117metadata$Moisture)

Water_FirstQ<-38.50
Water_Median<-42.16 
Water_ThirdQ<- 54.97

MMB117metadata$MoistureCAT<- cut(MMB117metadata$Moisture,c(0, 45,  84.68  ))

MMB117metadata$MoistureCAT  

levels(MMB117metadata$MoistureCAT )  
#[1] "(0,45]"    "(45,84.7]"

levels(MMB117metadata$MoistureCAT )  <- c("0 - 45 %","45 - 84.7 %")

MMB117metadata$MoistureCAT


ggplot(subset(MMB117metadata, Sample!="Neg. control"), aes(x=pH_Ca, y=ASV_divSha, colour=Site)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter() +
  scale_color_manual(values=c("springgreen4", "chocolate4","purple4","royalblue"))+
  facet_grid(Site ~ MoistureCAT) +
  theme_bw(base_size = 14)+
  theme(strip.background = element_rect(colour = "black", fill = "white"))+
  theme(panel.border = element_rect(colour = "black"))+
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank())+
  #guides(color=guide_legend(title="Sample type"))+
  theme(legend.position="none")

# Looks like in our case we don't have enough observations for this interaction
```